A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-08 podc_paper/algorithm-2.tex
working on GC section
removed propagate from helping in GC code

Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode3GC}.
The \op{Propagate} and \op{FindResponse} routines are unchanged.
Minor additions to \op{IndexDequeue} and \op{GetEnqueue} update the \var{last} field of nodes
visited by the routines to mark the blocks they visit in those nodes.
Only minor modifications have been made to \op{Enqueue}, \op{Append}, \op{CreateBlock}
to accomodate the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
of the \head\ field).
\op{Dequeue} has been similarly modified, and the second half of the routine is now in a
separate routine called \op{CompleteDequeue} so that it can also be used by other processes
helping to complete the operation.

Once a dequeue has been propagated to the root, 
\bf{\op CompleteDequeue} computes its answer.  This routine can also be called by helpers 
who want to help complete the dequeue.
In addition to calling \op{FindResponse} and \op{IndexDequeue} to compute the response,
it also updates \var{last} entries if the computed response is \nl.
It does this by calling \op{IndexDequeue} again, this time with \tr\ as its final parameter.
This extra parameter has been added to 
\op{IndexDequeue} to indicate whether it should mark blocks that it traverses (which all contain the dequeue).

@@

\Function{Object}{Dequeue()}{} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,\\
	    $\fld{index}\assign h$}\label{deqNewGC}
	}
    \State \Call{Append}{\var{B}}
    \gc{\State \Return \Call{CompleteDequeue}{\var{leaf}, $h$}}
\EndFunction{Dequeue}

@@

\gc{
\Function{Object}{CompleteDequeue}{\var{leaf}, $h$}
    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1, \fa}$
    \State $\var{res} \assign \Call{FindResponse}{b, i}$
	\If{$\var{res} = \nl$} \linecomment{update \var{last} values}
		\State $\Call{IndexDequeue}{\var{leaf}, h, 1, \tr}$
	\EndIf
	\State \Return{\var{res}}
\EndFunction{CompleteDequeue}
}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $p^2$}
    	\State \linecomment{Do garbage collection}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
		\EndFor
        \State \Call{Help}{}
        \State $T' \assign\ \Call{Split}{T,m-1}$
        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
        \State \Return \Call{Insert}{$T',B$}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}
\spac

\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{248}
\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
	\gc{
	\State $T \assign v.\fld{blocks}$\label{refreshReadTGC}
    \State \Call{SetSuper}{$v$, $T$}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$
		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}\label{casGC}
		}
    \EndIf

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
    \EndIf
\EndFunction{FindResponse}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
    	\State \Call{Propagate}{$\ell.\fld{parent}$}
        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$} 
            \State \linecomment{operation is a  dequeue}
            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDequeue}{\ell, \var{B}.\fld{index}}$ 
        \EndIf
    \EndFor
\EndFunction{Help}
}
\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{318}

@@

@@

\begin{figure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{340}

\end{algorithmic}
\caption{\label{pseudocode3GC}Bounded-space queue implementation, continued. }
\end{figure}

>>>>

Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode3GC}.
The \op{Propagate} and \op{FindResponse} routines are unchanged.
Minor additions to \op{IndexDequeue} and \op{GetEnqueue} update the \var{last} field of nodes
visited by the routines to mark the blocks they visit in those nodes.
Only minor modifications have been made to \op{Enqueue}, \op{Append}, \op{CreateBlock}
to accomodate the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
of the \head\ field).
\op{Dequeue} has been similarly modified, and the second half of the routine is now in a
separate routine called \op{CompleteDeq} so that it can also be used by other processes
helping to complete the operation.

Once a dequeue has been propagated to the root, 
\bf{\op CompleteDeq} computes its answer.  This routine can also be called by helpers 
who want to help complete the dequeue.
In addition to calling \op{FindResponse} and \op{IndexDequeue} to compute the response,
it also updates \var{last} entries if the computed response is \nl.
It does this by calling \op{IndexDequeue} again, this time with \tr\ as its final parameter.
This extra parameter has been added to 
\op{IndexDequeue} to indicate whether it should mark blocks that it traverses (which all contain the dequeue).

@@

\Function{Object}{Dequeue()}{} 
    \gc{
    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,\\
	    $\fld{index}\assign h$}\label{deqNewGC}
	}
    \State \Call{Append}{\var{B}}
    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}, $h$}}
\EndFunction{Dequeue}

@@

\gc{
\Function{Object}{CompleteDeq}{\var{leaf}, $h$}
    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1, \fa}$
    \State $\var{res} \assign \Call{FindResponse}{b, i}$
	\If{$\var{res} = \nl$} \linecomment{update \var{last} values}
		\State $\Call{IndexDequeue}{\var{leaf}, h, 1, \tr}$
	\EndIf
	\State \Return{\var{res}}
\EndFunction{CompleteDeq}
}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $p^2$}
    	\State \linecomment{Do garbage collection}\label{GCstart}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
		\EndFor
        \State \Call{Help}{}
        \State $T' \assign\ \Call{Split}{T,m-1}$
        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
        \State \Return \Call{Insert}{$T',B$}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}
\spac

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
	\gc{
	\State $T \assign v.\fld{blocks}$\label{refreshReadTGC}
    \State \Call{SetSuper}{$v$, $T$}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$
		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}\label{casGC}
		}
    \EndIf

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{248}
\Function{void}{Propagate}{\typ{Node} \var{v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
    \EndIf
    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
        \State \Call{Propagate}{\var{v}.\fld{parent}}
    \EndIf
\EndFunction{Propagate}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
    \EndIf
\EndFunction{FindResponse}

@@

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ 
        					\State \hspace*{4mm} and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
            \State \linecomment{operation is a propagated dequeue}
            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
        \EndIf
    \EndFor
\EndFunction{Help}
}

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
    \If {$\var{v} = \var{root}$} \Return{true}
    \ElsIf {$\var{v}.\fld{super} = \nl$ or 
    			\State \hspace*{4mm}$\var{v}.\fld{parent.blocks}[\var{v}.\fld{super}] = \nl$} 
		\State\Return{\fa}
    \Else
    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} 
				\State\var{sup} \assign\ $\var{sup}+1$
	    \EndIf
	    \State \hangbox{\Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and\\
	    		 $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and\\
			 	 \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}}
    \EndIf
\EndFunction{Propagated}
}

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{358}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{335}

\end{algorithmic}

@@

\begin{figure}

\caption{\label{pseudocode3GC}Bounded-space queue implementation, continued. }
\end{figure}

### 2023-01-08 podc_paper/correctness-2.tex
removed propagate from helping in GC code

Now we consider the additional steps a process takes to do GC in line \ref{GCstart}--\ref{GCend}.
\here{To be continued}

>>>>

Now we consider the additional steps a process takes to do GC in line \ref{GCstart}--\ref{GCend}.
For each integer $r$, each process does GC on each node $v$ with a block whose \fld{index} is $rp^2$ at most
once, since either its CAS succeeds, or another processes CAS advances the maximum 
Each such helping phase takes $O(p)$ time to read the \var{last} array,
$O(p \log p \log(p+q_{max}))$ time to \op{Help} (excluding any GC phases within \op{Help}, which will be accounted for at the other nodes),
and \here{to be continued}
\end{proof}
